Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-15 VMCAI-revised-version/final.tex
effectivity

We introduce five types of effectiveness \cite{BFM-ic17} that will be essential for solving various resilience problems in different models of WSTS.
\noindent
%			{\bf On effectivity.}
Let $\mathscr{S}=(S, \rightarrow, \leq)$ be a WSTS. We say that $\mathscr{S}$ is 
\begin{itemize}
\item {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
\item {\em post-effective} if $\mathscr{S}$ is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$. 
\item {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
\item {\em ideal-effective} \cite{BFM-ic17} if (1) the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable; (2) inclusion of ideals is decidable; (3) the downward closure $\downarrow \post(I)$ expressed as a finite union of ideals is computable from the ideal $I$.
\item {\em completion-post-effective} if $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ is post-effective, and there exists an algorithm $M_\downarrow$ that computes 
 $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
 $s_1, s_2, \ldots, s_m \in S$.
\end{itemize}
\alain{le rappel ci-dessous est-il entièrement utile ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Coverability is  decidable for a class of WSTS without the pred-basis hypothesis.

\begin{theorem} (Theorem 44 in \cite{BFM-ic17})
Coverability is  decidable for completion-post-effective WSTS.
\end{theorem}

@@

	\noindent	
	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
	which moreover begins
by writing the integer $j$ on its tape,	
	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
	The
	transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
	is a post-effective $\omega^2$-WSTS with strong compatibility
	and
	has the property that
	it is
	$\uparrow (1,1)$-resilient
	iff TM$_j$ halts on input $j$,
	since $(1,1)$ is coverable from $(0,0)$
	iff TM$_j$ halts on input $j$. 
	Hence there is no Turing machine which correctly determines 
	resilience and halts whenever its input is a
	WSTS.
Hence resilience for $\omega^2$-WSTS with strong compatibility is undecidable.\qed
\end{proof}

@@

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective $\omega^2$-WSTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}
\alain{a-t-on le même rsultat pour les WSTS avec pred* calculable ?}
\begin{proof}
Let us first recall two results in \cite{BFM-ic17}. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

This implies that resilience and bounded resilience are equivalent to coverability.
Since coverability is decidable for completion-post-effective $\omega^2$-WSTS, we deduce that both the 
  resilience problem and the bounded resilience problem are decidable.

@@

%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}. 
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for $\omega^2$-WSTS as well.\alain{déjà dit avant en mieux omega2 et strong comp et post-effective}\qed
\end{proof}

>>>>

We introduce five types of effectiveness \cite{BFM-ic17} that will be essential for solving various resilience problems in different models of WSTS.
\noindent
%			{\bf On effectivity.}
We say that a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ is 
\begin{itemize}
\item {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
\item {\em post-effective} if $\mathscr{S}$ is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$. 
\item {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
\item {\em ideally-effective} \cite{BFM-ic17} if (1) the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable; (2) inclusion of ideals is decidable; (3) the downward closure $\downarrow \post(I)$ expressed as a finite union of ideals is computable from the ideal $I$.
%
\item {\em completion-post-effective} if $\mathscr{S}$ is ideally-effective and its completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ is post-effective.
%
%and there exist an algorithm $M_\downarrow$ that computes 
% $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
% $s_1, s_2, \ldots, s_m \in S$.
%
\end{itemize}
\alain{le rappel ci-dessous est-il entièrement utile ? Je pense qu'on peut l'enlever, et toi ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Coverability is  decidable for a class of WSTS without the effective pred-basis hypothesis.

\begin{theorem} \cite{BFM-ic17}
Coverability is  decidable for completion-post-effective WSTS.
\end{theorem}

@@

	\noindent	
	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
	which moreover begins
by writing the integer $j$ on its tape,	
	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
	The
	transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
	is a post-effective $\omega^2$-WSTS with strong compatibility
	and
	has the property that
	it is
	$\uparrow (1,1)$-resilient
	iff TM$_j$ halts on input $j$,
	since $(1,1)$ is coverable from $(0,0)$
	iff TM$_j$ halts on input $j$. 
	Hence there is no Turing machine which correctly determines 
	resilience and halts whenever its input is a
	WSTS.
Hence resilience for $\omega^2$-WSTS with strong compatibility is undecidable.\qed
\end{proof}
 Let us remark that the transition system 
	$S_j = (\N^2, \{f_j, g\}, \leq)$ in the previous proof is not ideally effective neither completion-post-effective.

@@

%
\begin{theorem}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective WSTS with strong compatibility and a set $\Safe = \uparrow \Safe$.
{\sc Resilience}, {\sc Bounded resilience} 
and {\sc $k$-resilience} are decidable.
\end{theorem}
\alain{a-t-on le même rsultat pour les WSTS avec pred* calculable ?}
\begin{proof}
Let us first recall two results in \cite{BFM-ic17}. Proposition 30 establishes a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ and the runs of its completion $\hat{\mathscr{S}}$. It states that if $x \xrightarrow{k} y$ in $\mathscr{S}$ then for every ideal $I \supseteq \downarrow x$, there exists an ideal $J \supseteq \downarrow y$ such that $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$. Proposition 29 establishes that if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ then for every $y \in J$, there exists $x \in I$ and $y' \geq y$ such that $x \xrightarrow{k'} y'$ in $\mathscr{S}$. Moreover, if $\mathscr{S}$ has transitive compatibility then $k’ \geq k$; if $\mathscr{S}$ has strong compatibility then $k’ = k$.
%

@@

This implies that resilience and bounded resilience are equivalent to coverability.
Since coverability is decidable for completion-post-effective WSTS \cite{BFM-ic17}, we deduce that both the 
  resilience problem and the bounded resilience problem are decidable.

@@

%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}. 
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for $\omega^2$-WSTS as well.\alain{déjà dit avant en mieux omega2 et strong comp et post-effective mais reset-VASS apportent effective pred-basis....}\qed
\end{proof}
